---
title: "Project Proposal - Data"
author: "Adrian Tran, Harshitha Bachina, Hayden King, Josh Shneyder"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

# Linear Regression

## Loading in and Cleaning Data
```{r}
gdp = read.csv("~/Desktop/stat-331-71-PA6/gdppercapita_us_inflation_adjusted.csv")
life = read.csv("~/Desktop/stat-331-71-PA6/life_expectancy_years.csv")
library(tidyverse)
library(ggplot2)
library(broom)
gdp_clean <- gdp |>
  pivot_longer(cols = "X1959":"X2019",
               names_to = "Year",
               values_to = "GDP")

life_clean <- life |>
  pivot_longer(cols = "X1959":"X2019",
               names_to = "Year",
               values_to = "Life_Expectancy") |>
  select("country", "Year", "Life_Expectancy")

joined_data = merge(gdp_clean, life_clean, by = c("Year", "country")) |>
  mutate(
    Year = str_remove(Year, "X"),
    thousand = if_else(str_detect(GDP, "k"), T, F),
    GDP = as.double(str_remove(GDP, "k")),
    GDP = if_else(thousand, GDP * 1000, GDP),
    lGDP = log(GDP)
  ) |>
  select(-thousand)|>
  drop_na()
```

## Data Visualization

### Linear

```{r}
ggplot(data=joined_data, aes(x=GDP, y=Life_Expectancy)) + 
  geom_point() +
  labs(x="GDP per capita", y="", subtitle = "Life Expectancy", title = "Relationship between GDP per capita and Life Expectancy") + 
  geom_smooth(method="lm")
model = lm(Life_Expectancy ~ GDP, data=joined_data)
tidy(model)
summary(model)$r.squared
```
$$\widehat{Life Expectancy} = 63.28 + 0.0003135GDP + u$$
The linear model seems to be a poor fit for this data. The $R^2$ is rather low and the graph shows that the relationship is logarithmic rather than linear. The linear model predicts that if GDP per capita increases by 1 dollar, life expectancy is predicted to increase by 0.0003135 years.

### Logarithmic

```{r}
ggplot(data=joined_data, aes(x=lGDP, y=Life_Expectancy)) + 
  geom_point() +
  labs(x="Ln(GDP per capita)", y="", subtitle = "Life Expectancy", title = "Relationship between Ln(GDP per capita) and Life Expectancy") +
  geom_smooth(method="lm")
logmodel = lm(Life_Expectancy ~ lGDP, data=joined_data)
tidy(logmodel)
summary(logmodel)$r.squared
```
$$\widehat{Life Expectancy} = 22.84 + 5.312lGDP + u$$
The logarithmic transformation of GDP per capita makes a linear model fit quite well. The $R^2$ is moderate, which is pretty good considering how many factors can influence life expectancy across countries and across time. This model predicts that if GDP per capita increases by 1%, life expectancy increases by approximately 0.05312 years.

From here on out, we will examine only the relationship between $lGDP$ and $Life Expectancy$.

### Changes Over Time

```{r}
years = 1959:2019
lms = matrix(nrow = length(years), ncol = 2)
lms[,1] = years
for(y in years) {
  year_data = joined_data |>
    filter(Year == y)
  year_model = lm(Life_Expectancy ~ lGDP, data=year_data)
  lms[y-1958, 2] = summary(year_model)$r.squared
}
lms = data.frame(lms)
colnames(lms) = c("Year", "Rsquared")
ggplot(data=lms, aes(x=Year, y=Rsquared)) +
  geom_line() +
  ylim(0,1) +
  labs(y="", subtitle = "R^2", title = "R^2 of log(GDP per capita) and Life Expectancy over time")
```
The relationship is pretty stable over time and has been trending upward for the past decade. Outliers can be seen in the noticeable dips in the $R^2$. These low outliers could also be seen on the previous graphs.
```{r}
joined_data <- joined_data|>
  mutate(residuals = logmodel$residuals)
outliers <- joined_data |>
  filter(abs(residuals) > 20)
outliers
```
